Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-30 manuscript/manuscript.Rmd
changes in appendices and supplements

---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author: 
- 
output:
  pdf_document:
    keep_tex: true
geometry: margin=1in 
mainfont: Times New Roman
fontsize: 12pt
header-includes:
  \usepackage{setspace}\doublespacing
  \usepackage{endnotes}
  \renewcommand{\enotesize}{\normalsize}
  \let\footnote=\endnote
bibliography: kileybib.bib
csl: asa.csl
abstract: "Does personal culture---a person's attitudes, beliefs, values, and practices---change over the life course, or is it largely fixed by adulthood? Recent work has reached substantially different conclusions on this question. We argue that this disagreement is an unintended consequence of the \"tournament of models\" approach researchers have used. We introduce a new measure for quantifying the proportion of systematic variance in panel data attributable to intrapersonal change versus stable interpersonal differences. Applying this measure to 609 items from seven surveys in five countries, we find that although intrapersonal change is common, it is generally small in magnitude. We also demonstrate that this measure is useful for comparing social groups by showing that intrapersonal change is less common among U.S. college graduates than among those without a college degree. Overall, we believe this new measure will be a useful methodological tool and help reconcile an important theoretical debate."
---

@@

Supplemental Materials A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, the estimated values of $\omega$, and the proportion of residual variance for each question. Interpersonal differences are almost always the largest component of the total variance and tend to account for between 55 and 70 percent of total variance, while intrapersonal change is always the smallest, typically accounting for between 3 and 8 percent of total variance. Residual variance tends to account for between 22 and 37 percent of variance, though on several questions residual variance is greater than 50 percent. This might indicate survey items with low reliability or ones that capture genuinely rapid fluctuations.

@@

As a further test of this finding, we compared the $\omega$ values when using the full duration of a panel compared to when we dropped the final wave and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative effect of duration on $\omega$, we should see that same effect within questions. Results from this analysis are presented in Supplemental Materials C and contradict the coefficient from the regression model; with a few exceptions, we found that the longer we observe the same question, the higher $\omega$ is. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time. This supports the conclusion that the GSS shows less intrapersonal change because of its shorter duration.

@@

>>>>

---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author: 
- 
output:
  pdf_document:
    keep_tex: true
geometry: margin=1in 
mainfont: Times New Roman
fontsize: 12pt
header-includes:
  \usepackage{caption}
  \usepackage{setspace}\doublespacing
  \usepackage{endnotes}
  \renewcommand{\enotesize}{\normalsize}
  \let\footnote=\endnote
bibliography: kileybib.bib
csl: asa.csl
abstract: "Does personal culture---a person's attitudes, beliefs, values, and practices---change over the life course, or is it largely fixed by adulthood? Recent work has reached substantially different conclusions on this question. We argue that this disagreement is an unintended consequence of the \"tournament of models\" approach researchers have used. We introduce a new measure for quantifying the proportion of systematic variance in panel data attributable to intrapersonal change versus stable interpersonal differences. Applying this measure to 609 items from seven surveys in five countries, we find that although intrapersonal change is common, it is generally small in magnitude. We also demonstrate that this measure is useful for comparing social groups by showing that intrapersonal change is less common among U.S. college graduates than among those without a college degree. Overall, we believe this new measure will be a useful methodological tool and help reconcile an important theoretical debate."
---

@@

Appendix A shows the distribution of $V(D)$ and $V(C)$ across panels, and Supplemental Materials A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, the estimated values of $\omega$, and the proportion of residual variance for each question. Interpersonal differences are almost always the largest component of the total variance and tend to account for between 55 and 70 percent of total variance, while intrapersonal change is always the smallest, typically accounting for between 3 and 8 percent of total variance. Residual variance tends to account for between 22 and 37 percent of variance, though on several questions residual variance is greater than 50 percent. This might indicate survey items with low reliability or ones that capture genuinely rapid fluctuations.

@@

As a further test of this finding, we compared the $\omega$ values when using the full duration of a panel compared to when we dropped the final wave and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative effect of duration on $\omega$, we should see that same effect within questions. Results from this analysis are presented in Appendix B and contradict the coefficient from the regression model; with a few exceptions, we found that the longer we observe the same question, the higher $\omega$ is. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time. This supports the conclusion that the GSS shows less intrapersonal change because of its shorter duration.

@@

# Appendix

## Appendix A: The Distribution of Variation Components

Figure A1 shows the distribution of $V(D)$ and $V(C)$ across all panels.

```{=latex}
\begin{figure}[htp]
\begin{center}
\caption*{Figure A1: The Distribution of V(D) and V(C)}
```
```{r figA1, out.width = "350px"}
knitr::include_graphics("../figures/figure_A1.png")
```
```{=latex}
\end{center}
\end{figure}
```

## Appendix B: $\omega$ Values When the Last Wave Removed

Figure A2 presents the alternative strategy we used to understand the effects of duration on intraindividual change. First, we restricted the sample such that we have at least 4 waves of observations for all participants. This reduced the number of items to 249. We then fitted the Life Course Adaption models to this sample. Second, we removed the last wave from all the observations, and refit the same models once more. In the final step, we calculated the $\omega$ values for each item. The Figure C1 shows a scatterplot for these two sets of observations.

```{=latex}
\begin{figure}[tb]
\begin{center}
\caption*{Figure A2: Difference in $\omega$ Across Two Sample Specifications} 
```
```{r figA2, out.width = "400px"}
knitr::include_graphics("../figures/figure_A2.png")
```
```{=latex}
\end{center}
\end{figure}
```

### 2023-09-30 manuscript/manuscript.tex
changes in appendices and supplements

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing \usepackage{endnotes} \renewcommand{\enotesize}{\normalsize} \let\footnote=\endnote
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quantifying Stability and Change in Personal Culture Using Panel Data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Supplemental Materials A presents the estimated proportion of variance
attributable to interpersonal differences and intrapersonal change, the
estimated values of \(\omega\), and the proportion of residual variance
for each question. Interpersonal differences are almost always the
largest component of the total variance and tend to account for between
55 and 70 percent of total variance, while intrapersonal change is
always the smallest, typically accounting for between 3 and 8 percent of
total variance. Residual variance tends to account for between 22 and 37
percent of variance, though on several questions residual variance is
greater than 50 percent. This might indicate survey items with low
reliability or ones that capture genuinely rapid fluctuations.

@@

As a further test of this finding, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Supplemental Materials C and
contradict the coefficient from the regression model; with a few
exceptions, we found that the longer we observe the same question, the
higher \(\omega\) is. We interpret this combination of findings as
suggesting that the kinds of questions asked for a longer time period
tend to demonstrate less intrapersonal change than questions asked over
shorter periods, rather than a true function of time. This supports the
conclusion that the GSS shows less intrapersonal change because of its
shorter duration.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{caption} \usepackage{setspace}\doublespacing \usepackage{endnotes} \renewcommand{\enotesize}{\normalsize} \let\footnote=\endnote
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quantifying Stability and Change in Personal Culture Using Panel Data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Appendix A shows the distribution of \(V(D)\) and \(V(C)\) across
panels, and Supplemental Materials A presents the estimated proportion
of variance attributable to interpersonal differences and intrapersonal
change, the estimated values of \(\omega\), and the proportion of
residual variance for each question. Interpersonal differences are
almost always the largest component of the total variance and tend to
account for between 55 and 70 percent of total variance, while
intrapersonal change is always the smallest, typically accounting for
between 3 and 8 percent of total variance. Residual variance tends to
account for between 22 and 37 percent of variance, though on several
questions residual variance is greater than 50 percent. This might
indicate survey items with low reliability or ones that capture
genuinely rapid fluctuations.

@@

As a further test of this finding, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Appendix B and contradict
the coefficient from the regression model; with a few exceptions, we
found that the longer we observe the same question, the higher
\(\omega\) is. We interpret this combination of findings as suggesting
that the kinds of questions asked for a longer time period tend to
demonstrate less intrapersonal change than questions asked over shorter
periods, rather than a true function of time. This supports the
conclusion that the GSS shows less intrapersonal change because of its
shorter duration.

@@

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

\hypertarget{appendix-a-the-distribution-of-variation-components}{%
\subsection{Appendix A: The Distribution of Variation
Components}\label{appendix-a-the-distribution-of-variation-components}}

Figure A1 shows the distribution of \(V(D)\) and \(V(C)\) across all
panels.

\begin{figure}[htp]
\begin{center}
\caption*{Figure A1: The Distribution of V(D) and V(C)}

\includegraphics[width=350px]{../figures/figure_A1}

\end{center}
\end{figure}

\hypertarget{appendix-b-omega-values-when-the-last-wave-removed}{%
\subsection{\texorpdfstring{Appendix B: \(\omega\) Values When the Last
Wave
Removed}{Appendix B: \textbackslash omega Values When the Last Wave Removed}}\label{appendix-b-omega-values-when-the-last-wave-removed}}

Figure A2 presents the alternative strategy we used to understand the
effects of duration on intraindividual change. First, we restricted the
sample such that we have at least 4 waves of observations for all
participants. This reduced the number of items to 249. We then fitted
the Life Course Adaption models to this sample. Second, we removed the
last wave from all the observations, and refit the same models once
more. In the final step, we calculated the \(\omega\) values for each
item. The Figure C1 shows a scatterplot for these two sets of
observations.

\begin{figure}[tb]
\begin{center}
\caption*{Figure A2: Difference in $\omega$ Across Two Sample Specifications}

\includegraphics[width=400px]{../figures/figure_A2}

\end{center}
\end{figure}
